---
title: 'Team 3 / WVS: Exploratory Linear Model Plots'
author: "Michael Friendly"
date: "June 27, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory model-based plots

One data analysis strategy for an unknown data set is to produce a set of purely
exploratory graphs of just data, and see what turns up.  

An alternative is to use simple
models, mainly as a smoothing technique to get a **higher-level summary**, with the
understanding that your model may not be the best summary, or may be totally **wrong**.
But, you can **tune** the model (non-linear terms, add interactions, ...) and then the
plotting step(s) **remain the same** for the new model(s).

The caveats are:

* Numeric variables may actually be factors, not quantitative, but you forgot to use `as.factor()`
* You don't understand the  direction of coding: is `1` the highest or lowest?
* You forgot to account for, or analyse missing data, which, annoyingly was not coded as
`NA` in the `RData` file, but used negative values a la SAS or SPSS for various reasons why
data was missing.

But, this can give a jump start over purely EDA approaches, and allows me to show some things from
R packages that John Fox & I have developed:  `car`, `effects`, `heplots`.

```{r load_pkgs}
library(effects)
library(car)
library(heplots)
```

Load our semi-cleaned data file
```{r}
load("WVS.Rdata")
head(WVS)
```


## Fit some univariate linear models

We chose three possible variables to treat as responses:

```
"V10",  # happiness
"V11",  # health
"V23",  # life satisfaction
```

The possible predictors kept changing from one discussion to the next.  What I'm using here are
just:

```
"V57",  # marital_status
"V58",  # kids
"V238", # social class
"V240", # sex
"V248"  # education
```

Let's fit a simple additive linear model to each of these.  The strategy here is to:

* Fit  `mod <- lm()`
* Run `summary(mod)` and/or `car::Anova(mod)` to see test statistics, $R^2$, etc.
* Run `effects::plot(allEffects(mod))` to see a visual summary of the model predicted values
for **each** predictor, controlling for all other variables in the model.

Think of an *effect plot* as one kind of visual summary of a table of (partial regression) coefficients.

### Happiness

```{r wvs.mod1}
wvs.mod1 <- lm(happiness ~ marital + kids + social_class + sex + education, 
		data=WVS)
summary(wvs.mod1)
```
With such a huge sample size, everything is a significant effect.  What we want to understand
is roughly the magnitude and direction of the effects of each predictor on happiness in a relatively compact display.

Effect plot, using the short-hand default `plot(allEffects(mod))`.  Details of the screen size are
important here.  There are many, many graphic options to improve presentation, none used here.

```{r mod1-effplot, fig.width=6, fig.height=4}
plot(allEffects(wvs.mod1))
```

**NB**: 

* In this `.Rmd` file, I don't have as much control over the overall plot size, aspect ratio, 
and other details as I would in a hand-crafted graph I re-sized on-screen or otherwise
mainpulated.

* This plot (by default) can be misleading, because each panel is scaled *separately*,
so small absolute effects for a given predictor can appearlarger than if a common scale was
used for all panels.

* This "model" looks pretty good, until we notice that the $R^2 =0.04$ !!! But--- hey, we *can* say
with a straight face that $R^2 > 0$, however,  We aren't using this for inference.



### Health

```{r}
wvs.mod2 <- lm(health ~ marital + kids + social_class + sex + education, data=WVS)
summary(wvs.mod2)

```

Effect plot:
```{r mod2-effplot, fig.width=6, fig.height=4}
plot(allEffects(wvs.mod2))
```

### Life satisfaction

```{r}
wvs.mod3 <- lm(life_sat ~ marital + kids + social_class + sex + education, data=WVS)
summary(wvs.mod3)
```

Effect plot:
```{r mod3-effplot, fig.width=6, fig.height=4}
plot(allEffects(wvs.mod3))
```


