---
title: "Data Journalism"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(knitr)
opts_chunk$set(echo = FALSE)
options(digits = 3)
```

# Data Journalism Lecture Quiz


These non-graded questions are designed to test your understanding of the materials presented in the [Data Journalism lecture](https://friendly.github.io/6135/lectures/DataJournalism.pdf).
They cover a broad range of topics from the presentation, including:

- Core components and techniques of data journalism
- Historical examples and landmark projects
- Key figures who advanced the field
- Tools and technologies used by data journalists
- Modern storytelling methods and emerging approaches

(This was actually an experiment. I asked [Claude 3.7 Sonnet](https://www.anthropic.com/claude/sonnet) to read the lecture slides and come up with 10 questions.)

## Question 1

```{r q1, echo=FALSE}
question("What are the **three main aspects** of Data Driven Journalism (DDJ) as described in the presentation?",
  answer("Crowdsourcing, satellite imagery, and dynamic updating"),
  answer("Database analysis, visualization, and software tools", correct = TRUE),
  answer("Election reporting, health statistics, and social media"),
  answer("FOI requests, investigative reporting, and infographics")
)
```

## Question 2

```{r q2, echo=FALSE}
question("Which data journalism project won the Pulitzer Prize in investigative reporting in 1989 for exposing racial discrimination in lending practices?",
  answer("The Panama Papers"),
  answer("The Color of Money", correct = TRUE),
  answer("Hurricane Andrew"),
  answer("The _Toronto Star_ racial profiling investigation")
)
```

## Question 3

```{r q3, echo=FALSE}
question("What _significant advancement_ in election reporting was introduced with CNN's Magic Wall in 2008?",
  answer("The ability to predict winners with 99% certainty"),
  answer("A large touch-screen display with interactive visualization of election data", correct = TRUE),
  answer("The \"election needle\" showing predicted outcomes"),
  answer("The use of hierarchical Bayes modeling for results")
)
```

## Question 4

```{r q4, echo=FALSE}
question("According to the 2022 data journalism survey mentioned in the presentation, what are the most commonly used tools for data analysis and visualization among journalists?",
  answer("Python and R"),
  answer("Excel and Google Sheets", correct = TRUE),
  answer("Tableau and D3.js"),
  answer("Datawrapper and Flourish")
)
```

## Question 5

```{r q5, echo=FALSE}
question("Which of the following is described as a modern technique for interactive storytelling in digital journalism?",
  answer("Before-After Circle Charts"),
  answer("Hierarchical Bayes modeling"),
  answer("Scrollytelling", correct = TRUE),
  answer("Satellite journalism")
)
```

## Question 6

```{r q6, echo=FALSE}
question("What significant investigation involved the analysis of 2 terabytes of data, exposing tax avoidance schemes across 210,000 companies in 21 offshore jurisdictions?",
  answer("WikiLeaks"),
  answer("The Panama Papers", correct = TRUE),
  answer("The _Toronto Star_ Trump Lies database"),
  answer("Our World in Data")
)
```

## Question 7

```{r q7, echo=FALSE}
question("Which technique did _The Guardian_ employ to analyze 700,000 documents detailing expense claims of British MPs?",
  answer("Satellite imaging"),
  answer("Scrollytelling"),
  answer("Crowd sourcing", correct = TRUE),
  answer("Visual explainers")
)
```

## Question 8

```{r q8, echo=FALSE}
question("What made the _Miami Herald_'s investigation of Hurricane Andrew's destruction in 1992 a landmark in data journalism?",
  answer("It was the first to use satellite imagery"),
  answer("It used crowd sourcing to gather damage reports"),
  answer("It combined investigative reporting, analysis, and detailed maps to show why homes were destroyed", correct = TRUE),
  answer("It used dynamic updating of storm tracking data")
)
```

## Question 9

```{r q9, echo=FALSE}
question("\"Borrowing strength\" in election prediction, which later became known as \"hierarchical Bayes,\" was pioneered by which statistician working with NBC in 1962?",
  answer("John Tukey", correct = TRUE),
  answer("Bill Dedman"),
  answer("Daniel Dale"),
  answer("Michael Friendly")
)
```

## Question 10

```{r q10, echo=FALSE}
question("What emerging field of journalism uses imagery from privately owned satellites to provide evidence for investigations?",
  answer("Crowd-sourced journalism"),
  answer("Scrollytelling"),
  answer("Satellite journalism", correct = TRUE),
  answer("Visual explainers")
)
```

