---
title: "Visualizing Uncertainty"
output: learnr::tutorial
runtime: shiny_prerendered
---
  
```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```

## Visualizing Uncertainty Quiz

This quiz is designed to test your understanding of concepts and techniques related to visualizing uncertainty in data, based on the lecture "Visualizing Uncertainty." The quiz covers a range of important topics, including:
  
- The nature and sources of uncertainty in statistics and visualization
- Different methods for visualizing distributions (histograms, density plots, boxplots, etc.)
- Advanced visualization techniques like violin plots, raincloud plots, and hypothetical outcome plots
- Error bars and confidence intervals (frequentist vs. Bayesian approaches)
- Uncertainty in fitted curves and models
- Cartographic uncertainty and geographic visualization techniques
- Best practices for conveying uncertainty in data analysis and presentation

Understanding how to properly visualize uncertainty is critical for effective data communication and preventing misinterpretation of statistical results.

## Question 1

```{r q1, echo=FALSE}
question("What are the **four main sources** of uncertainty in statistics according to the lecture?",
         answer("Sampling, measurement, modeling, and visualization"),
         answer("Data, assumptions, models, and replications", correct = TRUE),
         answer("Noise, variance, bias, and standard error"),
         answer("Error bars, p-values, confidence intervals, and credible intervals"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         message = "At the highest level, uncertainty of results arises in data (sampling variability, assumptions required for
                 the validity of results, the form of models used for analysis and the potential for replications to
                 yield different results."
)
```

## Question 2

```{r q2, echo=FALSE}
question("What is a key problem with boxplots for visualizing distributions?",
         answer("They always require a large sample size"),
         answer("They can only show categorical data"),
         answer("They are insensitive to multi-modal data", correct = TRUE),
         answer("They cannot display outliers effectively"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         message = "Because boxplots summarize the data by quartiles, they are insensitive to multi-modal data."
)
```

## Question 3

```{r q3, echo=FALSE}
question("What is the difference between standard deviation and standard error?",
         answer("Standard deviation measures variability in the data while standard error measures uncertainty about the mean", correct = TRUE),
         answer("Standard deviation is used for categorical data while standard error is used for continuous data"),
         answer("Standard deviation is a Bayesian concept while standard error is a frequentist concept"),
         answer("Standard deviation is always larger than standard error regardless of sample size"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         message = "Standard deviation and standard error are both frequentist ideas used for continuous data, but the
            standard deviation measures variability in the data while standard error measures uncertainty about the mean."
)
```

## Question 4

```{r q4, echo=FALSE}
question("In the context of kernel density estimation, what does the bandwidth parameter control?",
         answer("The number of observations included in the analysis"),
         answer("The width of the moving window used for smoothing the distribution", correct = TRUE),
         answer("The height of the resulting density curve"),
         answer("The statistical significance of the distribution"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Question 5

```{r q5, echo=FALSE}
question("What is a 'raincloud plot' in data visualization?",
         answer("A plot that shows precipitation patterns over time"),
         answer("A combination of density curve, boxplot, and jittered points", correct = TRUE),
         answer("A method for visualizing uncertainty in weather forecasts"),
         answer("A specialized chart used only for time-series data"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Question 6

```{r q6, echo=FALSE}
question("What is the correct interpretation of a frequentist 95% confidence interval?",
         answer("There is a 95% probability that the true parameter falls within this interval"),
         answer("If we take repeated samples, 95% of the resulting intervals will contain the true parameter", correct = TRUE),
         answer("95% of the data points fall within this interval"),
         answer("The model explains 95% of the variation in the data"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Question 7

```{r q7, echo=FALSE}
question("How do Hypothetical Outcome Plots (HOPs) convey uncertainty?",
         answer("By showing all possible outcomes simultaneously"),
         answer("By visualizing a set of draws from a distribution either as small multiples or animation", correct = TRUE),
         answer("By providing numerical probabilities for each possible outcome"),
         answer("By using color intensity to represent probability"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Question 8

```{r q8, echo=FALSE}
question("What was a historical method for indicating uncertainty in maps before modern techniques?",
         answer("Using monsters and imagined creatures to mark unexplored areas", correct = TRUE),
         answer("Leaving uncertain areas completely blank"),
         answer("Using only black and white for unknown regions"),
         answer("Including warning text directly on the map"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         message ="Cartographers like Abraham Ortelius use fanciful drawings of monsters and imagninary creatures to
         show that there might be something at a given location, but he didn't know what."
)
```

## Question 9

```{r q9, echo=FALSE}
question("In a meta-analysis, what does a funnel plot help to identify?",
         answer("The overall effect size"),
         answer("Which studies have the most precise measurements"),
         answer("Publication bias", correct = TRUE),
         answer("The sample size required for future studies"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         message = "The funnel plot is a plot of effect size on the _X_ axis against a measure of standard error (inverse precision)
             on the _Y_ axis. Points outside a central triangle indicate potential publication bias."
)
```

## Question 10

```{r q10, echo=FALSE}
question("According to the {ggdist} package design, what are the three key elements that can be combined for visualizing uncertainty?",
         answer("Mean, median, and mode"),
         answer("Data, distribution, and interval", correct = TRUE),
         answer("Points, lines, and bars"),
         answer("Color, shape, and size"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Quiz Complete

```{r results='asis'}
random_praise()
```


Congratulations on completing the quiz on Human Factors in Data Visualization!
  

```{r context="server"}
shiny::observeEvent(
  input$get_score, 
  {
    objs2 = learnr:::get_tutorial_state()
    
    # Number of correct questions
    
    n_correct <- 
      # Access the $correct sublist item in each list item
        lapply(objs2, purrr::pluck, "correct") |>
           # make it a vector containing: TRUE and FALSE and NAs
           # NA is appearing for list items which don't have
           # a $correct subitem
                unlist() |> 
           # Taking the sum of a logical Vector returns the number of TRUEs
                sum(na.rm=TRUE)
    
    # Number of total questions
    
    total_questions <- 
      # 1. Access $type in each list item and make it a vector of types
      lapply(objs2, purrr::pluck, "type") |> unlist()
    
    # 2. Count the number of "question" in that vector
    total_questions <- total_questions[total_questions == "question"] |> 
      length()
      
      
    output$score = shiny::renderText(
      paste0(n_correct, " out of ", total_questions,
        " question(s) were correct.")
)
    invisible()
  }
)
```

```{r score, echo=FALSE}
shiny::actionButton("get_score", "Get Score", icon=icon("star"))
shiny::br()
shiny::br()
shiny::textOutput("score")
```

You can review your answers and explanations by navigating back through the questions.
  